Missing the Point\(\colon\) Non-Convergence in Iterative Imputation Procedures

### 2020-04-17 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Minor changes in methods section

Main question: How can non-convergence be diagnosed? Sub-questions: Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used? How many iterations are sufficient/needed to reach convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? What are the effects of continued iterations on estimates, predictions and inferences? And finally, do the answers differ with varying missingness proportions?

@@

```
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  for (number of simulation runs from 1 to 1000)
    2. create missingness
    for (number of iterations from 1 to 100)
      3. impute missingness
      4. compute convergence diagnostics
      5. perform analysis of scientific interest
      6. pool results across imputations
      7. compute performance measures
  8. aggregate outcomes 
9. combine outcomes
```
The aim of the simulation study is to evaluate the impact of: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2, ..., 100$), and 2) varying the difficulty of the missingness problem (**different missingness proportions** $prop = .05, .25, .50, .75, .95$). For the purpose of this study, sampling variance is not of interest. Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd].

The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study). A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`, where

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} \sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\!\!\text{, and}\\[2\jot]
Y &=  1 + 2X_1 + .5X_2 - X_3 + \epsilon .
\end{align*} 
Subsequently, 
We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely  The quantity of scientific interest is the estimated regression coeffiecent estimand is the A finite population of $N=1000$ is simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):

$$Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$

The data generating model is a multivariate normal distribution

Outcome variable $Y$ is subsequently calculated as 
$$Y =  1 + 2X_1 + .5X_2 - X_3 + \epsilon .$$
<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $y$ [@rubin87].  -->

**(Add the function `mvtnorm::rmvnorm()` back in somewhere here? If yes, use citation("mvtnorm").)**

>>>>

Main question: How can non-convergence be diagnosed? Sub-questions: Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? **(These are part of the methods, not RQ per se: What are the effects of continued iterations on estimates, predictions and inferences? And finally, do the answers differ with varying missingness proportions? That is, we vary the nr of iterarions and the missingness proportion because we assume that the algorithm has not reached convergence at $t=1$, and performs worse with moderate to high missingness (not so much with little or a LOT of missingness).)**

@@

```
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  for (number of simulation runs from 1 to 1000)
    2. create missingness
    for (number of iterations from 1 to 100)
      3. impute missingness
      4. compute convergence diagnostics
      5. perform analysis of scientific interest
      6. pool results across imputations
      7. compute performance measures
  8. aggregate outcomes across simulation runs 
9. combine outcomes of all missingness proportions
```
The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2, ..., 100$), and 2) varying the missingness proportions ($p_{miss} = .05, .25, .50, .75, .95$). The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints. **As the number of iterations goes up, the imputation chains will become independent of the inital values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem. The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data. Higher missingness proportions then cause slower convergence. However, if the fraction of missing information is very high, there is so little information in the data that the random component in the algorithm will take the overhand and a stable but very uncertain (high variance) point will be reached.

The data-generating mechanism is a multivariate normal distribution, representing person data on three independent variables (from an unspecified social scientific field of study). Let

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim  \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} 
For the purpose of this study, sampling variance is not of interest. Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd]. A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`. Subsequently, a fourth variable is constructed to serve as dependent variable in a multiple linear regression problem. Let 
$$
Y_i =  1 + 2X_{1i} + .5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $i = 1, 2, ..., N$ and $\epsilon \sim \mathcal{N}(0, 100)$.

We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely  ... solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):

$$Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.$$
<!-- This note only considers a 'missing completely at random' (MCAR) mechanism, where the probability of being missing is equal for all $n \times p$ cells in $y$ [@rubin87].  -->

### 2020-04-17 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Minor changes in methods section

Main question: How can non-convergence be diagnosed? Sub-questions: Are
common MCMC non-convergence diagnostics appropriate for MICE? And if so,
which threshold should be used? How many iterations are
sufficient/needed to reach convergence? Are the default number of
iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
What are the effects of continued iterations on estimates, predictions
and inferences? And finally, do the answers differ with varying
missingness proportions?

@@

\begin{verbatim}
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  for (number of simulation runs from 1 to 1000)
    2. create missingness
    for (number of iterations from 1 to 100)
      3. impute missingness
      4. compute convergence diagnostics
      5. perform analysis of scientific interest
      6. pool results across imputations
      7. compute performance measures
  8. aggregate outcomes 
9. combine outcomes
\end{verbatim}

The aim of the simulation study is to evaluate the impact of: 1)
terminating the MICE algorithm at different imputation chain lengths
(\(t = 1, 2, ..., 100\)), and 2) varying the difficulty of the
missingness problem (\textbf{different missingness proportions}
\(prop = .05, .25, .50, .75, .95\)). For the purpose of this study,
sampling variance is not of interest. Therefore, a single complete set
may serve as comparative truth in all simulation runs \citep{vinknd}.

The data-generating mechanism is a multivariate normal distribution. We
simulate a finite population of \(N=1000\) to represent person data for
an unspecified social scientific field of study.

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3\\
\epsilon
\end{pmatrix} &\sim  N
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5\\
0
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0\\
0 & 0 & 0 & 100
\end{pmatrix}
\end{bmatrix}\!\!\text{, and}\\[2\jot]
Y &=  1 + 2X_1 + .5X_2 - X_3 + \epsilon .
\end{align*}

A finite population of \(N=1000\) is simulated to solve a multiple
linear regression problem, where dependent variable \(Y\) is regressed
on independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
notation with betas here, and add what the quantity/-ies of scientific
interest is/are!}):

@@

The data generating model is a multivariate normal distribution

Outcome variable \(Y\) is subsequently calculated as
\[Y =  1 + 2X_1 + .5X_2 - X_3 + \epsilon .\]

\textbf{(Add the function \texttt{mvtnorm::rmvnorm()} back in somewhere
here? If yes, use citation(``mvtnorm'').)}

>>>>

Main question: How can non-convergence be diagnosed? Sub-questions: Are
common MCMC non-convergence diagnostics appropriate for MICE? And if so,
which threshold should be used to diagnose non-convergence? How many
iterations are sufficient/needed to be able to diagnose non-convergence?
Are the default number of iterations sufficient (i.e., 5 in mice, 10 in
SPSS and Stata, 30 in mi)? \textbf{(These are part of the methods, not
RQ per se: What are the effects of continued iterations on estimates,
predictions and inferences? And finally, do the answers differ with
varying missingness proportions? That is, we vary the nr of iterarions
and the missingness proportion because we assume that the algorithm has
not reached convergence at \(t=1\), and performs worse with moderate to
high missingness (not so much with little or a LOT of missingness).)}

@@

\begin{verbatim}
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  for (number of simulation runs from 1 to 1000)
    2. create missingness
    for (number of iterations from 1 to 100)
      3. impute missingness
      4. compute convergence diagnostics
      5. perform analysis of scientific interest
      6. pool results across imputations
      7. compute performance measures
  8. aggregate outcomes across simulation runs 
9. combine outcomes of all missingness proportions
\end{verbatim}

The aim of the simulation study is to evaluate the impact of inducing
non-convergence by: 1) terminating the MICE algorithm at different
imputation chain lengths (\(t = 1, 2, ..., 100\)), and 2) varying the
missingness proportions (\(p_{miss} = .05, .25, .50, .75, .95\)). The
assumption underlying the different number of iterations is that the
algorithm generally does reach convergence at \(t=1\), because the
initial values are sampled randomly from the set of observed datapoints.
\textbf{As the number of iterations goes up, the imputation chains will
become independent of the inital values until the point at which an
extra added iteration does not lead to a more converged state.} The
second set of experimental conditions under consideration--the
missingness proportions--are chosen to reflect the difficulty of the
missingness problem. The inherent assumption is that low missingness
proportions will lead to quick algorithmic convergence, since there is a
lot of information in the observed data. Higher missingness proportions
then cause slower convergence. However, if the fraction of missing
information is very high, there is so little information in the data
that the random component in the algorithm will take the overhand and a
stable but very uncertain (high variance) point will be reached.

The data-generating mechanism is a multivariate normal distribution,
representing person data on three independent variables (from an
unspecified social scientific field of study). Let

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim  \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} For the purpose of this study, sampling variance is not of
interest. Therefore, a single complete set may serve as comparative
truth in all simulation runs \citep{vinknd}. A finite population of
\(N=1000\) is simulated using the function \texttt{mvtnorm::rmvnorm()}.
Subsequently, a fourth variable is constructed to serve as dependent
variable in a multiple linear regression problem. Let \[
Y_i =  1 + 2X_{1i} + .5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2, ..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

We consider four quantities of scientific interest \citep[`conceptual
estimands';][]{morr19}, namely \ldots{} solve a multiple linear
regression problem, where dependent variable \(Y\) is regressed on
independent variables \(X_1\), \(X_2\) and \(X_3\) (\textbf{check
notation with betas here, and add what the quantity/-ies of scientific
interest is/are!}):

@@
